---
title: "Make a Plot"
author: "Wycliffe"
format:
  html:
    embed-resources: true
editor: visual
execute:
  warning: false
editor_options:
  chunk_output_type: console
---

```{r}
library(usethis)
use_git_config(user.name = "wotieno", user.email = "wycliffeawino246@gmail.com")
```

## Task 1: Import

The required packages for this homework exercise have already been added to the following code-chunk.

1.  Run the following code chunk. Tipp: Click on the green play button in the top right corner of the code chunk.

```{r}
library(gapminder)
library(ggplot2)
library(dplyr)
```

## Task 2: Transform for data in 2007

**Read and understand the code**

1.  Read the code below out loud, verbally describing what each line does.

2.  Run the code chunk below.

3.  Observe the output in the console.

```{r}
gapminder |>
  filter(year == 2007)
```

4.  Write in your own words what you think this code does:

*Your answer here: The code chunk above reads the dataframe gapminder and transforms it by filtering for instances where the year is 2007.*

5.  Now run the code chunk below that assigns the result to an object.

6.  **Tip:** Watch the Environment tab in RStudio (top right pane) to see the new object appear.

```{r}
gapminder_2007 <- gapminder |>
  filter(year == 2007)
```

7.  Click on `gapminder_2007` in the Environment tab to view the data. What do you notice about this dataset compared to the original `gapminder` dataset?

*Your answer here: The code chunk above reads the dataframe gapminder, transforms it for cases where the year is 2007 and stores the new dataset in an object called gapminder_2007.*

## Task 3: Summarize data for life expectancy by continent

**Read and understand the code**

1.  Read the code below out loud, verbally describing what each line does.

2.  Run the code chunk below.

3.  Observe the output in the console.

```{r}
gapminder_2007 |>
  group_by(continent) |>
  summarise(
    count = n(),
    lifeExp = median(lifeExp)
  )
```

4.  Write in your own words what you think this code does:

*Your answer here: This code chunk takes the object gapminder_2007 (which is a dataframe) and transform it further by grouping observations by the continent. Next, it summarises the observations by creating a new variable called 'count', which is the number of countries represented in the corresponding continent and returns the median life expectancy in that group.*

5.  Now run the code chunk below that assigns the result to an object.

6.  **Tip:** Watch the Environment tab to see how this new object differs from `gapminder_2007`.

```{r}
gapminder_summary_2007 <- gapminder_2007 |>
  group_by(continent) |>
  summarise(
    count = n(),
    lifeExp = median(lifeExp)
  )
```

7.  Look at the `gapminder_summary_2007` object in the Environment tab. How many rows and columns does it have? Why is this different from `gapminder_2007`?

*Your answer here: `gapminder_summary_2007` has 5 rows and 3 columns. It shows the number of continents in the dataset (=5), so we only have that number of groups. The 3 columns represents the name of the grouping variable –continent– and two others that we specify in our summary –lifeExp & count. `gapminder_2007` on the other hand is the original dataframe filtered for the year 2007.*

## Task 4: Summarize data for life expectancy by continent and year

**Read and understand the code**

1.  Read the code below out loud, verbally describing what each line does.

2.  Run the code chunk below.

3.  Observe the output in the console.

```{r}
gapminder |>
  group_by(continent, year) |>
  summarise(lifeExp = median(lifeExp))
```

4.  Write in your own words what you think this code does:

*Your answer here: This code chunk takes the original dataframe gapminder and transform it through the pipe operator by grouping by continent and years then summarizing the by the median values for life expectancy in that continent in that year.*

5.  Now run the code chunk below that assigns the result to an object.\$

6.  **Tip:** Notice how the Environment tab shows different information for different types of objects.

```{r}
gapminder_summary_continent_year <- gapminder |>
  group_by(continent, year) |>
  summarise(lifeExp = median(lifeExp))
```

7.  Compare this result with Task 3. What's different about grouping by two variables instead of one? How does this affect the output?

*Your answer here: First, we do not create the variable count by which we were summarizing the results. Next, for each continent, it returns the corresponding year and summarizes the results by the median life expectancy in that year.*

## Task 5: Create your first plot

Now that you understand how to transform data, let's create a plot.

1.  Read the code below out loud, verbally describing what you think it will do.

2.  Run the code chunk.

```{r}
ggplot(data = gapminder_2007, 
       mapping = aes(x = continent, 
                     y = lifeExp)) +
  geom_boxplot()
```

3.  Write in your own words what this plot shows:

*Your answer here: This is a boxplot showing the distribution of life expectancy by continent. Africa has the widest distribution and lowest mean life expectancy with mean \~55.5 years, followed by Asia with a mean of \~73.0 years. Americas is third , followed by Europe. Oceania has the lowest distribution and the highest mean life expectancy \~81.0 years.*

## Task 6: Data communication

**In the YAML header (between the three dashes at the top of the document)**

1.  Add your name as the author of this document.

2.  Render the document and fix any errors.

## Task 7: Commit & Push & Issue

-   Refer back to the instructions for Step 2 and Step 3 on the course website to complete the assignment 02a: <https://ds4owd-002.github.io/website/content/assignments/md-02/am-02-1-data-visualisation.html#step-2-commit-and-push-your-changes>
